% Functional Programming in R
% David Springate
% University of Manchester Centre for Biostatistics
% @datajujitsu
% May 2013


What is functional programming?
===============================

Programming metaphysics
-----------------------

* Programs are representations of reality on the computer
* There are different ways to represent reality...

***************

OOP / imperative metaphysics
----------------------------

* C, Python, Java etc.
* Everything is an object with state and behaviour

### e.g. A river is a "thing" with various attributes:

* flow rate 
* depth
* pollution levels

***************

Functional Metaphysics
----------------------

"No man ever steps in the same river twice, for it's not the same river and he's not the same man." - Heraclitus


* R, Lisp, Haskell
* Focus is on processes rather than objects
* Things are a collection of fixed values which go through processes (functions) over time
* Values are fixed (immutable)

### FUNCTIONAL RIVER DIAGRAM

*****************

Elements of FP
--------------

* Functions are : 
    - Closures - Functions that return functions ("Function Factories")
        - Wrappers
    - Higher order functions - Functions that take functions as arguments ("Functionals")
        - For loop replacements
* Vectorised / declarative expressions
* Recursion

************

Advantages of FP
----------------

* Easier debugging
* Self-documenting code
* Clear and concise code 
* Parallelisation
* No more difficult

************

Functional programming in R
===========================

R parentage
-----------

C/Fortran -> S  
Lisp -> Scheme 
S + Scheme -> R

So you can write R that looks like C - UGLY

or

R that looks like Lisp - BEAUTIFUL

FP is often discovered later and is seen as confusing

In R everything is a function!

****************



Declarative programming / vectorised operations
------------------------------------------------
```{r}
# C style vector allocation:
system.time({
x <- c()
for(i in 1:200000){
    if(i %% 2 == 0){
        x <- c(x, i)
    }
}
})

# Takes ~ 11 seconds
# FP style:

system.time({
a <- 1:200000
x <- a[a %% 2 == 0]
})

# Takes ~ 
```

************

Recursion
---------

```{r}

a <- list(1, 2, 3, 4, list(5, 6, 7, 8), 9, list(10, list(11, 12)))

tree_count <- function(l){
    if(is.null(l[[1]])) {
        return(0)
    } else if(is.list(l[[1]])){
        return(tree_count(l[[1]]) + tree_count(l[2:length(l)]))
    } else {
        return(l[[1]] + tree_count(l[2:length(l)]))
    } 
}
tree_count(a)

sum(unlist(a))
```

*************

Higher-order functions
----------------------

Functions that operate on all elements of a collection (vector/list/dataframe)

apply : apply a function to rows or columns of a matrix/dataframe


lapply : apply a function to a collection, return a list 
vector in, list out

sapply : apply a function to a collection, return a vector/matrix

vector in, vector out

Reduce : collection in, scalar out


****************

Further reading
---------------

